Tropical Mathematics and the Lambda Calculus

### 2023-01-23 section1.tex
commit

%This is not a \emph{ad-hoc} setting: 
In the last decades, tropical geometry evolved into a vast and rich research domain, providing a combinatorial counterpart of usual algebraic geometry, with important connections with optimisation theory \cite{Sturmfelds}.
Computationally speaking, working with $\min$ and $+$ is generally easier than working with standard addition and multiplication; for instance, the fundamental (and generally intractable) problem of finding the roots of a polynomial admits a \emph{linear time} algorithm in the tropical case (moreover, the tropical roots of a polynomial $p$ can be used to approximate the actual roots of $p$ \cite{Noferini2015}).
The computational nature of tropical notions explains why these are so widely applied in computer science, notably for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).

Coming back to our previous discussion on program semantics, tropical geometry might seem to provide precisely what we are looking for, as the monomials $x^{n}$ turn precisely into the Lipschitz functions $nx$.
At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail. Furthermore, connections between tropical linear algebra and metric spaces (in the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}), have also been observed \cite{Fuji}.

>>>>

%This is not a \emph{ad-hoc} setting: 
In the last decades, tropical geometry evolved into a vast and rich research domain, providing a combinatorial counterpart of usual algebraic geometry, with important connections with optimisation theory \cite{Sturmfelds}.
Computationally speaking, working with $\min$ and $+$ is generally easier than working with standard addition and multiplication; for instance, the fundamental (and generally intractable) problem of finding the roots of a polynomial admits a \emph{linear time} algorithm in the tropical case (and, moreover,  the tropical roots can be used to approximate the actual roots \cite{Noferini2015}).
The computational nature of tropical notions explains why these are so widely applied in computer science, notably for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).

Coming back to our previous discussion on program semantics, tropical geometry might seem to provide precisely what we are looking for, as it turns the monomials $x^{n}$ into the Lipschitz functions $n\cdot x$.
At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail. Furthermore, connections between tropical linear algebra and metric spaces (in the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}), have also been observed \cite{Fuji}.

### 2023-01-23 section4.tex
commit

A tropical polynomial is a piece-wise linear function $f:\Lawv\to \Lawv$ of the form \begin{equation}
f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}
\end{equation}
where the $i_{j}$ are natural numbers and the $c_{i_{j}}$ are in $\Lawv$.
For example, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$
are illustrated in Fig.~\ref{fig:plot1} for $n=2,3,4$.
A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable. In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes). For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.

@@

\begin{figure}
\begin{subfigure}{0.24\textwidth}
\begin{tikzpicture}[scale=0.55]
\begin{axis}[samples=250]
\addplot[red,domain=0:0.8] {min(2*x+1/4, x+1/2, 1)};
\addplot[blue,domain=0:0.8] {min(3*x+1/8,2*x+1/4, x+1/2, 1)-0.01};
\addplot[orange,domain=0:0.8] {min(4*x+1/16,3*x+1/8,2*x+1/4, x+1/2, 1)-0.02};

@@

\end{axis}
\end{tikzpicture}
\caption{}
\label{fig:plot2}
\end{subfigure}
\label{fig:plot12}
\caption{
%\ref{fig:plot1}
 (a) Plot of the tropical polynomials $\varphi_{2}$ (in red), $\varphi_{3}$ (in blue), $\varphi_{4}$ (in orange) and of their limit tLs $\varphi$ (in violet). The points where the slope changes are  the tropical roots of $\varphi$, i.e.~the points $x=2^{-(i+1)}$, satisfying $ix+2^{-i}=(i+1)x+2^{-(i+1)}$.
%\ref{fig:plot2}
(b) Plot of
$\varphi(x,y)=\min\{2x, 2x+y,3y\}$.}
\end{figure}

>>>>

A tropical polynomial is a piece-wise linear function $f:\Lawv\to \Lawv$ of the form \begin{equation}
f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}
\end{equation}
where the $i_{j}$ are natural numbers and the $c_{i_{j}}$ are in $\Lawv$.
For example, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$
are illustrated in Fig.~\ref{fig:plot1} for $n\leq 4$.
A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable. In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes). For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.

@@

\begin{figure}
\begin{subfigure}{0.24\textwidth}
\begin{tikzpicture}[scale=0.55]
\begin{axis}[samples=250]
\addplot[yellow,domain=0:0.8] {1+0.02};

\addplot[orange,domain=0:0.8] {min(x+1/2, 1)+0.01};

\addplot[red,domain=0:0.8] {min(2*x+1/4, x+1/2, 1};
\addplot[blue,domain=0:0.8] {min(3*x+1/8,2*x+1/4, x+1/2, 1)-0.01};
\addplot[orange,domain=0:0.8] {min(4*x+1/16,3*x+1/8,2*x+1/4, x+1/2, 1)-0.02};

@@

\end{axis}
\end{tikzpicture}
\caption{}
\label{fig:plot2}
\end{subfigure}
\label{fig:plot12}
\caption{
%\ref{fig:plot1}
 (a) Plot of the tropical polynomials $\varphi_{n}$, for $n\leq 4$ (from top to bottom), and of their limit tLs $\varphi$ (in violet). The points where the slope changes are  the tropical roots of $\varphi$, i.e.~the points $x=2^{-(i+1)}$, satisfying $ix+2^{-i}=(i+1)x+2^{-(i+1)}$.
%\ref{fig:plot2}
(b) Plot of
$\varphi(x,y)=\min\{2x, 2x+y,3y\}$.}
\end{figure}
